Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-12-12 doc/manuscript/translocation.qmd
Minor text clean-up

<!--# Revise to reflect reordering of sections -->

Vast alteration of the biosphere by humans is causing a sixth mass extinction.
Adaptation to modified environments often is the only means for species to persist, but the extent to which such "evolutionary rescue" can prevent or reverse biodiversity loss is largely unknown.
Using results from genomic analyses, a 15-year reintroduction effort, and population modeling, we provide a compelling example of evolution reversing the declines of an imperiled amphibian and allowing the reestablishment of extirpated populations.
The once-common mountain yellow-legged (MYL) frog is threatened with extinction by the human-facilitated emergence of a lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd").
Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually begin to recover.
Using an exome capture approach, we show that MYL frogs have undergone substantial evolutionary change following disease outbreaks.
Moreover, in naturally recovering populations observed changes in immune function loci are associated with increased resistance/tolerance to Bd infection.
Large-scale reintroduction of frogs from rescued populations resulted in the establishment of reproducing populations despite ongoing disease.
In addition, results from viability modeling suggest that established populations have a low probability of extinction over 50 years.
Collectively, these results provide a rare example of how evolutionary rescue and the reintroduction of resistant/tolerant individuals can allow the landscape-scale recovery of disease-impacted species.
This example has potentially broad implications for the many taxa worldwide that are threatened with extinction by anthropogenic stressors.

@@

<!--# Revise to reflect reduced emphasis on evolutionary rescue. -->

Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
One possibility is rapid evolution that restores positive growth to declining populations ("evolutionary rescue").
Individuals from these rescued populations could also be used to reestablish extinct populations.
We evaluated both possibilities in a once-common frog species driven to the brink of extinction by a disease that has devastated global amphibian biodiversity and for which few options exist to reverse disease impacts.
We found that frogs have evolved increased resistance to infection, facilitating population recovery.
In addition, introduction of frogs from these populations allowed reestablishment despite the ongoing presence of disease.
Such reintroductions could play an important role in biodiversity conservation in our rapidly changing world.

@@

<!--# Remove uncited references in .bib -->

@@

<!--# Mark: Add brief description here of methods used in comparison of survival by translocated vs. naturally recruited individuals. Primary focus should be on describing the fact that data were derived from mrmr model results, and on how those data were extracted. -->

@@

Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals <!--# At the beginning of this section, we state that 53 individuals were sampled. I assume the difference reflects the fact that some individuals were excluded by the filtering process? Clarify where 50 came from.  -->using VCFtools [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
Additionally, we calculated per-locality $\pi$ using 100kb sliding windows with VCFtools and compared these values using a boxplot and Wilcoxon rank-sum tests.
Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.

>>>>

Vast alteration of the biosphere by humans is causing a sixth mass extinction, driven in part by an increase in emerging infectious diseases.
The emergence of the lethal fungal pathogen (*Batrachochytrium dendrobatidis*; "Bd") has devastated global amphibian biodiversity, with hundreds of species experiencing declines or extinctions.
Given that there are no broadly applicable methods to reverse these impacts in the wild, the future of many amphibians appears grim.
The once-common mountain yellow-legged (MYL) frog is emblematic of amphibians threatened by Bd.
Although most MYL frog populations are extirpated following disease outbreaks, some persist and eventually recover.
Frogs in these recovering populations have increased resistance against Bd infection, consistent with evolution of resistant genotypes and/or acquired immunity.
We conducted a 15-year landscape-scale reintroduction study and show that frogs collected from recovering populations and reintroduced to vacant habitats can reestablish populations despite the presence of Bd.
In addition, results from viability modeling suggest that many reintroduced populations have a low probability of extinction over 50 years.
To better understand the role of evolution in frog resistance, we compared the genomes of MYL frogs from Bd-naive and recovering populations.
We show that frogs have undergone substantial evolutionary change in response to Bd exposure and that changes in immune function loci are consistent with increased resistance.
These results provide a rare example of how the reintroduction of resistant individuals can allow the landscape-scale recovery of disease-impacted species.
This example has broad implications for the many taxa worldwide that are threatened with extinction by disease.

@@

Understanding how species persist despite accelerating global change is critical for the conservation of biodiversity.
Emerging infectious diseases can have particularly devastating impacts, and few options typically exist to reverse these effects.
We used large-scale reintroductions of disease-resistant individuals in an effort to recover a once-common frog species driven to the brink of extinction by a disease that has recently devastated global amphibian biodiversity.
Introduction of resistant frogs allowed reestablishment of populations in the presence of disease.
In addition, resistance appears to have a genetic basis and to have evolved recently following exposure of frogs to Bd.
The evolution of resistance and reintroductions of resistant individuals could play an important role in biodiversity conservation in our rapidly changing world.

@@

@@

@@

Using the liberal set of outlier variants, we performed a statistical overrepresentation test (Fisher's exact test) using Blast2GO [@conesa2005].
First, we retrieved the BLAST hits and mapped GO terms for each gene in our targeted transcriptome.
Next, we compared the 30 unique outlier genes from the liberal set of outlier variants to the complete set of genes in our target transcriptome to determine if any GO biological functions, molecular functions, or cellular processes were overrepresented.
We then repeated this process for the set of 33 genes located in the overlapping *F~ST~* and $\pi_{diff}$ splined windows.
To compare global patterns of genetic diversity between naive and persistent populations, we calculated individual heterozygosity for all 50 individuals <!--# At the beginning of this section, we state that 53 individuals were sampled. I assume the difference reflects the fact that some individuals were excluded by the filtering process? Clarify where 50 comes from.  -->using VCFtools [@danecek2011], displayed these values using a violin plot, and conducted a Wilcoxon rank-sum test for significant differences between groups.
Additionally, we calculated per-locality $\pi$ using 100kb sliding windows with VCFtools and compared these values using a boxplot and Wilcoxon rank-sum tests.
Code used for genomic analyses and to create figures is available from the following GitHub repository: <https://github.com/allie128/mylf-selection>.
